---
title: "Auditing donor influence in pseudobulk analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Auditing donor influence in pseudobulk analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4)
```

This package asks how a **fixed two-condition pseudobulk analysis** changes
when one biological donor is removed. A large effect change is descriptive:
it can reflect real donor heterogeneity, limited precision, or other data
properties. The package does not decide that a donor should be excluded.

## Construct an auditable example

The simulated counts below are only a workflow example, not a benchmark or
biological result. Each column of the `SummarizedExperiment` is one observed
sample and cell type. Donor `d4` lacks a `B`-cell pseudobulk under `stim`.
The full sample registry makes that absence visible.

```{r input}
set.seed(1103)
sample_table <- data.frame(
    sample_id = paste0("s", seq_len(8)),
    donor_id = rep(paste0("d", seq_len(4)), each = 2),
    condition = rep(c("ctrl", "stim"), 4))
source_rows <- rbind(
    transform(sample_table, cell_type = "T"),
    transform(sample_table[seq_len(7), ], cell_type = "B"))
counts <- matrix(rnbinom(250 * nrow(source_rows), mu = 25, size = 5),
                 nrow = 250,
                 dimnames = list(paste0("g", seq_len(250)),
                                 paste0("column", seq_len(nrow(source_rows)))))
source_rows$n_cells <- 30L
se <- SummarizedExperiment::SummarizedExperiment(
    assays = list(counts = counts),
    colData = S4Vectors::DataFrame(source_rows))
pb <- scDonorAudit::preparePseudobulk(
    se, "sample_id", "donor_id", "condition", "cell_type",
    sample_table = sample_table, n_cells = "n_cells")
contrast <- c(numerator = "stim", denominator = "ctrl")
```

The named contrast fixes the sign: positive log2 fold-change means higher
expression in `stim` than `ctrl`. The count assay must contain raw,
non-negative integer counts. The package checks numeric validity but cannot
prove the assay was not normalized upstream.

## Review the design before fitting

The default paired policy is strict. An incomplete eligible pair blocks that
cell type without affecting other cell types.

```{r audit}
audit <- scDonorAudit::auditDesign(pb, "paired", contrast)
as.data.frame(audit[["spec"]])
subset(as.data.frame(audit[["sample_membership"]]),
       cell_type == "B" & reason_code != "OK")
```

`plan_status = "not_created"` means a deletion plan was never established for
the blocked cell type. It does not mean that all donor deletions succeeded or
that the result was stable.

## Fit and inspect the ledger

```{r fit}
result <- scDonorAudit::assessDonorInfluence(pb, "paired", contrast)
subset(as.data.frame(result[["fits"]]),
       select = c(cell_type, run_id, omitted_donor,
                  execution_status, reason_code))
head(as.data.frame(scDonorAudit::summarizeInfluence(result,
                                                     effect_threshold = 0.5))[,
    c("cell_type", "gene_id", "n_planned", "n_effect_valid",
      "max_abs_delta_observed")])
```

The summary's `n_planned` includes all eligible donor deletions, even failed
refits. A maximum effect change is only the maximum among refits with valid
effects. Missing effects are never converted into zero changes or
non-significant tests. Adjusted p values use a fixed baseline gene family
within each cell type; they do not control error across cell types or deletion
runs.

To analyze the `B`-cell data from complete pairs, request the alternative
policy explicitly. This changes the analysis cohort, so report the excluded
pair when presenting the result.

```{r complete-pairs}
reduced <- scDonorAudit::assessDonorInfluence(
    pb, "paired", contrast, pair_policy = "complete_pairs")
subset(as.data.frame(reduced[["spec"]]),
       select = c(cell_type, plan_status, n_planned, n_samples))
```

## Inspect figures

The coverage plot shows included and excluded sample/cell-type combinations.
Gene plots require explicit, stable gene IDs; they do not choose genes by
apparent influence.

```{r coverage-plot}
scDonorAudit::plotInfluence(result, "coverage")
```

```{r gene-plots}
scDonorAudit::plotInfluence(result, "influence", cell_type = "T",
                            gene_id = c("g1", "g2", "g3"))
scDonorAudit::plotInfluence(result, "effect", cell_type = "T", gene_id = "g1")
```

Before interpreting a gene, inspect `result[["fits"]]`,
`result[["sample_membership"]]`, and the number of valid deletion runs.
A failed fit or unestimable contrast limits the observed sensitivity analysis.

## Reproduce a real single-nucleus example

The package includes a compact pseudobulk derived from the public
[Crowell19 mouse cortex dataset](https://doi.org/10.6084/m9.figshare.8976473.v1)
(ExperimentHub `EH3297`). Raw counts from 1,740 Astrocyte nuclei were summed
within each of eight mice, retaining all 11,076 genes. Four mice received LPS
and four received Vehicle. The original single-nucleus object is available
through ExperimentHub and is not bundled here. The included
`extdata/README-crowell19.md` records the transformation, attribution, and
source's CC BY 4.0 license.

```{r real-data}
count_file <- system.file("extdata", "crowell19_astrocyte_counts.csv",
                          package = "scDonorAudit")
sample_file <- system.file("extdata", "crowell19_astrocyte_samples.csv",
                           package = "scDonorAudit")
real_counts <- as.matrix(utils::read.csv(count_file, row.names = 1,
    check.names = FALSE))
real_samples <- utils::read.csv(sample_file)
real_source <- SummarizedExperiment::SummarizedExperiment(
    assays = list(counts = real_counts),
    colData = S4Vectors::DataFrame(real_samples))
real_registry <- real_samples[, c("sample_id", "donor_id", "condition")]
real_pb <- scDonorAudit::preparePseudobulk(real_source, "sample_id",
    "donor_id", "condition", "cell_type", sample_table = real_registry,
    n_cells = "n_cells")
real_contrast <- c(numerator = "LPS", denominator = "Vehicle")
real_audit <- scDonorAudit::auditDesign(real_pb, "independent",
    real_contrast, min_cells = 20L)
as.data.frame(real_audit[["spec"]])
real_result <- scDonorAudit::assessDonorInfluence(real_pb, "independent",
    real_contrast, min_cells = 20L)
table(as.data.frame(real_result[["fits"]])$execution_status)
head(as.data.frame(scDonorAudit::summarizeInfluence(real_result))[,
    c("gene_id", "n_planned", "n_effect_valid",
      "max_abs_delta_observed")])
```

This example demonstrates the data and audit workflow. Eight mice offer
limited replication, especially after one is omitted. A large observed effect
change can reflect biological heterogeneity or imprecision; it does not label
a mouse for exclusion. The p values are those of the specified edgeR model,
not a test that a donor is aberrant.

## Session information

```{r session-info}
sessionInfo()
```
