---
title: "Auditing donor influence in pseudobulk analysis"
author: "Fuhao Jiang"
date: "2026-09-30"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Auditing donor influence in pseudobulk analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4)
```

## Introduction

This package asks how a **fixed two-condition pseudobulk analysis** changes
when one biological donor is removed. A large effect change is descriptive:
it can reflect real donor heterogeneity, limited precision, or other data
properties. The package does not decide that a donor should be excluded.

### Related work and scope

[scuttle](https://bioconductor.org/packages/scuttle) provides cell aggregation.
The following tools and analyses overlap with different parts of this workflow:

| Tool or work | Primary task | Relationship to this package |
|:--|:--|:--|
| [muscat](https://bioconductor.org/packages/muscat) and [scran `pseudoBulkDGE()`](https://bioconductor.org/packages/scran) | Multi-sample or cell-type-wise differential expression | Supply established pseudobulk analysis approaches; the audit here concerns changes across a planned set of whole-donor omissions. |
| [dreamlet](https://bioconductor.org/packages/dreamlet) | Pseudobulk analysis for complex designs | Already reports assay- and gene-level errors; this package adds a narrower two-condition deletion plan and its effect summary. |
| [scFastDE](https://bioconductor.org/packages/devel/bioc/html/scFastDE.html) | Donor-weighted pseudobulk differential expression, including paired designs | Weighting and deletion sensitivity answer different questions. |
| [SampleQC](https://github.com/wmacnair/SampleQC) | Multi-sample single-cell quality control | Quality metrics and changes to a specified DE contrast are distinct outputs. |
| [Study-specific LODO pseudobulk analysis](https://pmc.ncbi.nlm.nih.gov/articles/PMC12991174/) | Refit after excluding each donor | Shows whole-donor deletion is not a new statistical idea. |

These approaches overlap in inputs or aims. scDonorAudit uses
[edgeR](https://bioconductor.org/packages/edgeR) for a specified two-condition
contrast and adds a consistent audit of the planned donor omissions, design
viability, execution events, and observed gene-effect changes. It is not a new
differential-expression test or a donor-outlier classifier.

The source repository is
[Emr-27/scDonorAudit](https://github.com/Emr-27/scDonorAudit).

## Installation

After Bioconductor accepts and releases the package, its standard installation
command will be:

```{r bioconductor-installation, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("scDonorAudit")
library(scDonorAudit)
```

For the current source version, install from a checked-out package directory
after installing the dependencies listed in `DESCRIPTION`:

```{r source-installation, eval=FALSE}
install.packages(".", repos = NULL, type = "source")
library(scDonorAudit)
```

## Construct an auditable example

The simulated counts below are only a workflow example, not a benchmark or
biological result. Each column of the `SummarizedExperiment` is one observed
sample and cell type. Donor `d4` lacks a `B`-cell pseudobulk under `stim`.
The full sample registry makes that absence visible.

```{r input}
set.seed(1103)
sample_table <- data.frame(
    sample_id = paste0("s", seq_len(8)),
    donor_id = rep(paste0("d", seq_len(4)), each = 2),
    condition = rep(c("ctrl", "stim"), 4))
source_rows <- rbind(
    transform(sample_table, cell_type = "T"),
    transform(sample_table[seq_len(7), ], cell_type = "B"))
counts <- matrix(rnbinom(250 * nrow(source_rows), mu = 25, size = 5),
                 nrow = 250,
                 dimnames = list(paste0("g", seq_len(250)),
                                 paste0("column", seq_len(nrow(source_rows)))))
source_rows$n_cells <- 30L
se <- SummarizedExperiment::SummarizedExperiment(
    assays = list(counts = counts),
    colData = S4Vectors::DataFrame(source_rows))
pb <- scDonorAudit::preparePseudobulk(
    se, "sample_id", "donor_id", "condition", "cell_type",
    sample_table = sample_table, n_cells = "n_cells")
contrast <- c(numerator = "stim", denominator = "ctrl")
```

The named contrast fixes the sign: positive log2 fold-change means higher
expression in `stim` than `ctrl`. The count assay must contain raw,
non-negative integer counts. The package checks numeric validity but cannot
prove the assay was not normalized upstream. Factor coding is fixed inside
the design, so a user's global `options("contrasts")` does not change the
named effect's direction or scale.

## Review the design before fitting

The default paired policy is strict. An incomplete eligible pair blocks that
cell type without affecting other cell types.

```{r audit}
audit <- scDonorAudit::auditDesign(pb, "paired", contrast)
as.data.frame(audit[["spec"]])
subset(as.data.frame(audit[["sample_membership"]]),
       cell_type == "B" & reason_code != "OK")
```

`plan_status = "not_created"` means a deletion plan was never established for
the blocked cell type. It does not mean that all donor deletions succeeded or
that the result was stable.

## Fit and inspect the ledger

```{r fit}
result <- scDonorAudit::assessDonorInfluence(pb, "paired", contrast)
subset(as.data.frame(result[["fits"]]),
       select = c(cell_type, run_id, omitted_donor,
                  execution_status, reason_code))
head(as.data.frame(scDonorAudit::summarizeInfluence(result,
                                                     effect_threshold = 0.5))[,
    c("cell_type", "gene_id", "n_planned", "n_effect_valid",
      "max_abs_delta_observed")])
```

The summary's `n_planned` includes all eligible donor deletions, even failed
refits. A maximum effect change is only the maximum among refits with valid
effects. Missing effects are never converted into zero changes or
non-significant tests. Adjusted p values use a fixed baseline gene family
within each cell type; they do not control error across cell types or deletion
runs.

The analysis first fixes the eligible samples and complete-pair policy, then
plans all whole-donor deletions. Genes retained by the full-data filter remain
the comparison family in each refit. The named contrast and covariate meanings
also stay fixed. Each successful deletion refit **re-estimates** library-size
normalization and the edgeR model on its remaining samples. Thus the analysis
reference is fixed, while fitted parameters are not.

Read each gene in this order: is its baseline effect valid; how many deletions
were planned; how many effects are comparable; then how large were the observed
changes? `n_sign_reversal` counts even tiny crossings of zero. With a positive
`effect_threshold`, `n_material_reversal` requires both effects to cross that
threshold in opposite directions, and `baseline_near_zero` flags a baseline
for which a raw sign change is especially easy to overread. The minimum and
maximum deletion effects are an **observed range**, not a confidence interval.
An unavailable effect may reflect design failure, backend failure, or a
gene-level status; it is not evidence of stability.

### Output dictionary

| Component | Meaning |
|:--|:--|
| `spec` | Per-cell-type plan status and number of planned donor omissions. |
| `sample_membership` / `coverage` | Registered sample-by-cell-type combinations and the actual baseline inclusion decisions. |
| `fits` / `fit_samples` | Baseline and deletion ledger. `design_reason_code` preserves the original design diagnosis; `execution_status`, `reason_code`, and `stage` describe the later execution outcome. The separate `fit_samples` table aligns sample IDs with library sizes and normalization factors for each completed run. |
| `gene_filter` / `results` | Fixed baseline gene family and gene-by-run effects, tests, validity flags, and gene status codes. |
| `issues` | Design problems and execution warnings or errors; a run can have both a backend warning and an error. |
| `config` / `provenance` | Requested design, contrast and thresholds; source preparation, package/R/edgeR versions, and the actual named contrast. |

Design rank deficiency can coexist with an estimable target contrast. A rank
problem is still a reason to skip fitting and is not relabeled as target
non-estimability. When a baseline cannot run, the planned deletion rows remain
in `fits`; `design_reason_code` keeps their independent audit results even if
their execution outcome becomes `BASELINE_UNAVAILABLE`.

To analyze the `B`-cell data from complete pairs, request the alternative
policy explicitly. This changes the analysis cohort, so report the excluded
pair when presenting the result.

```{r complete-pairs}
reduced <- scDonorAudit::assessDonorInfluence(
    pb, "paired", contrast, pair_policy = "complete_pairs")
subset(as.data.frame(reduced[["spec"]]),
       select = c(cell_type, plan_status, n_planned, n_samples))
```

## Inspect figures

The coverage plot shows included and excluded sample/cell-type combinations.
Gene plots require explicit, stable gene IDs; they do not choose genes by
apparent influence.

```{r coverage-plot}
scDonorAudit::plotInfluence(result, "coverage")
```

```{r gene-plots}
scDonorAudit::plotInfluence(result, "influence", cell_type = "T",
                            gene_id = c("g1", "g2", "g3"))
scDonorAudit::plotInfluence(result, "effect", cell_type = "T", gene_id = "g1")
```

Before interpreting a gene, inspect `result[["fits"]]`,
`result[["sample_membership"]]`, and the number of valid deletion runs.
A failed fit or unestimable contrast limits the observed sensitivity analysis.
Grey influence tiles mean the effect is unavailable; consult the fit ledger
and gene status for the specific cause.

## Reproduce a real single-nucleus example

The package includes a compact pseudobulk derived from the public
[Crowell19 mouse cortex dataset](https://doi.org/10.6084/m9.figshare.8976473.v1)
(ExperimentHub `EH3297`). Raw counts from 1,740 Astrocyte nuclei were summed
within each of eight mice, retaining all 11,076 genes. Four mice received LPS
and four received Vehicle. The original single-nucleus object is available
through ExperimentHub and is not bundled here. The included
`extdata/README-crowell19.md` records the transformation, attribution, and
source's CC BY 4.0 license.

```{r real-data}
count_file <- system.file("extdata", "crowell19_astrocyte_counts.csv",
                          package = "scDonorAudit")
sample_file <- system.file("extdata", "crowell19_astrocyte_samples.csv",
                           package = "scDonorAudit")
real_counts <- as.matrix(utils::read.csv(count_file, row.names = 1,
    check.names = FALSE))
real_samples <- utils::read.csv(sample_file)
real_source <- SummarizedExperiment::SummarizedExperiment(
    assays = list(counts = real_counts),
    colData = S4Vectors::DataFrame(real_samples))
real_registry <- real_samples[, c("sample_id", "donor_id", "condition")]
real_pb <- scDonorAudit::preparePseudobulk(real_source, "sample_id",
    "donor_id", "condition", "cell_type", sample_table = real_registry,
    n_cells = "n_cells")
real_contrast <- c(numerator = "LPS", denominator = "Vehicle")
real_audit <- scDonorAudit::auditDesign(real_pb, "independent",
    real_contrast, min_cells = 20L)
as.data.frame(real_audit[["spec"]])
knitr::kable(as.data.frame(real_audit[["sample_membership"]])[, c(
    "sample_id", "donor_id", "condition", "n_cells", "in_baseline")])
real_result <- scDonorAudit::assessDonorInfluence(real_pb, "independent",
    real_contrast, min_cells = 20L)
table(as.data.frame(real_result[["fits"]])$execution_status)
knitr::kable(as.data.frame(real_result[["fits"]])[, c(
    "run_id", "omitted_donor", "execution_status",
    "design_reason_code", "reason_code")])
head(as.data.frame(scDonorAudit::summarizeInfluence(real_result))[,
    c("gene_id", "n_planned", "n_effect_valid",
      "max_abs_delta_observed")])
```

This example demonstrates the data and audit workflow. Eight mice offer
limited replication, especially after one is omitted. A large observed effect
change can reflect biological heterogeneity or imprecision; it does not label
a mouse for exclusion. The p values are those of the specified edgeR model,
not a test that a donor is aberrant.

## Read the donor-deletion results

The fit ledger above has one full-data fit and eight donor-deletion fits. To
illustrate what the summary adds, consider genes with valid effects in all
eight deletions and an absolute full-data log2 fold-change of at least 1.
Order these genes by their largest observed absolute change, breaking ties by
gene ID, and show the genes at the 10th and 90th percentile ranks. This rule
selects examples with different sensitivity; it is not a performance test or
a biological gene-selection procedure.

```{r real-interpretation-selection}
real_summary <- as.data.frame(scDonorAudit::summarizeInfluence(real_result))
candidates <- real_summary[
    real_summary$complete_effect_coverage &
    is.finite(real_summary$baseline_logFC) &
    abs(real_summary$baseline_logFC) >= 1 &
    is.finite(real_summary$max_abs_delta_observed), ]
candidates <- candidates[order(candidates$max_abs_delta_observed,
                               candidates$gene_id), ]
chosen <- candidates[c(ceiling(0.1 * nrow(candidates)),
                       ceiling(0.9 * nrow(candidates))), ]
chosen$illustration <- c("lower change", "higher change")
knitr::kable(chosen[, c("illustration", "gene_id", "baseline_logFC",
    "n_planned", "n_effect_valid", "max_abs_delta_observed")], digits = 3)
```

For the higher-change example, inspect every deletion instead of relying on
one maximum. A missing or failed deletion would appear in the fit ledger and
reduce `n_effect_valid`; it must not be interpreted as zero change.

```{r real-interpretation-refits}
real_se <- real_result[["results"]][[chosen$cell_type[2L]]]
gene_row <- match(chosen$gene_id[2L], rownames(real_se))
refits <- as.data.frame(SummarizedExperiment::colData(real_se))
refits$logFC <- as.numeric(SummarizedExperiment::assay(real_se,
    "logFC")[gene_row, ])
refits$change_from_full <- refits$logFC - refits$logFC[1L]
largest <- which.max(abs(refits$change_from_full[-1L])) + 1L
knitr::kable(refits[, c("run_id", "omitted_donor", "logFC",
                         "change_from_full")], digits = 3)
```

In this dataset, `r nrow(candidates)` genes meet the stated selection rule.
For `r chosen$gene_id[2L]`, the full-data LPS-versus-Vehicle log2 fold-change
is `r sprintf("%.3f", chosen$baseline_logFC[2L])`; omitting
`r refits$omitted_donor[largest]` changes it by
`r sprintf("%.3f", refits$change_from_full[largest])` to
`r sprintf("%.3f", refits$logFC[largest])`. All
`r chosen$n_effect_valid[2L]` planned deletion effects are valid here, and
the direction remains negative. By comparison, the lower-change example
`r chosen$gene_id[1L]` has a maximum absolute change of
`r sprintf("%.3f", chosen$max_abs_delta_observed[1L])`. These observations
describe how sensitive the specified model is to donor omission in this
dataset. They do not show that any mouse is erroneous, that a gene is robust
in another dataset, or that the selected genes are representative of all
genes.

## Session information

```{r session-info}
sessionInfo()
```
